|

### 2023-06-28 talks/ISF_talk/talk.qmd
post ISF

---
title: "alloscore, ISF 43"
format:
  revealjs: 
    width: 1500
revealjs-plugins:
  - attribution
editor: visual
css: styles.css
---

@@

![](hub_timeline.png){}

@@

@@

## The Role of IDF Hubs

@@

- with trend toward elicitation of probabilistic forecasts

@@

But in the inimitable words of chatGPT:

> It's important to note that while these efforts represent significant advances in the field, forecasting the spread and impact of infectious diseases is a very complex task with many challenges, and these models should be interpreted with caution.

::: notes
Over last two decades, multicenter effort to collect forecasts via hubs for ID outbreaks (epi/pan/endemic) with central purpose to "inform public health responses"

With uncertainty quantification being well-established as needed DM input,
:::

## The fundamental need to decide allocations

Basic element of any such response is an **allocation decision** regarding

::: {.column width="10%"}

<div style="height: 200px; width: 50px;">
  <svg height="100%" width="100%" viewBox="0 0 10 20">
    <line x1="5" y1="0" x2="5" y2="15" style="stroke:black;stroke-width:2" />
    <polygon points="0,20 10,20 5,15" style="fill:black;" />
  </svg>
</div>

::: {.column width="40%"}

-   medical supplies
-   personnel
-   facility capacity
-   R&D funding
-   public tolerance for NPIs
:::

::: {.column width="50%"}
Similar to other contexts where there are large-scale institutional efforts to use expert judgment to assess and manage social risk

-   finance
-   meteorology
-   military conflict
:::
:::

::: notes
Allocations can be wrt space, time, demographic

Even with complex game-theoretic approachs to collective decision making, allocation almost always a requirement to strategy development
:::

@@

$S$ if **proper** if $\displaystyle{E_{F}[S(F,Y)] \leq E_{G}S(F,Y)]}$ for all $F, G$

@@

Hubs have tried to use SRs to rank and combine forecasts in ways that offer best suppport for PH response to ID outbreaks

Primary scoring rule used currently is the **Weighted Interval Score** which is a discrete approximation to the 
**Continuous Ranked Probability Score**

@@

@@

::: {.important}

@@

Given $l$, we have an *automatically proper* scoring rule $S(F,y) := l(x^F, y)$.

::: {.callout-note appearance="simple" icon=false}

### Our running example:

@@

@@

$$
\begin{align}
l(x,y) &= \sum_{i=1}^{52} \max(y_i - x_i, 0) \\
x^F &\in \left\{ \text{possible } x' \,\middle| \, E_{F}[l(x',Y)] \leq E_{F}[l(x,Y)] \text{ for any possible } x \right\}
\end{align}
$$

Novelty of our approach for ID forecasting is how we define $x$ being 
**possible**

 - as satisfying some resource constraint $\displaystyle{\sum x_i \leq K}$

::: {.callout-note appearance="simple" icon=false}

### Note:

@@

@@

::: {.callout-note appearance="Important" icon=false}

## A central goal in design of scoring rules for social good

Using an explicit vacabulary of social loss in ID forecasting hub activities has instrinsic advantage
over use of "off-the-shelf" statistical functionals from statistic inference

@@

@@

$$ 
\begin{align}
L(x, \lambda ; K, F) &= \sum_{i=1}^N E_{F_i}\left[l(x_i, Y_i)]\right)+\lambda\left(\sum_{i=1}^N x_i-K\right) \\  
\nabla_{x} L = 0 &\implies 
E_F \left[\frac{d}{dx_i}(Y_i - x_i)_{+}\right] + \lambda 
= F_i(x_i)-1 + \lambda = 0 \\
&\implies x_i = F_i^{-1}(1-\lambda)
\end{align}
$$

Solve $\sum_{l=1}^{N} F_{i}^{-1}(1-\lambda) = K$ to get $x_{i}^F = F_i^{-1}(1-\lambda^{\star})$

Even for simple parametric $F_i$, this rarely has an analytic solution.

Given quantile forecasts $q_{i,j}, j = 1,\ldots,23$, we interpolate full $F_i$ to form $L(x, \lambda ; K, F)$

Adopt an iterative Lagrange multiplier method following long history in inventory management of solving **constrained multi-product newsvendor** problems

illustrate algo with a lambda over xs plot against tau

___

![](plots/state_hosps_ens_only.jpeg)

## Scoring against an oracle

To get more interpretable allocation scores we adjust by the score of an oracle that allocates *fairly* based on what demands actually occur

Note: The oracle is not the DGP, it is a time-traveller

Comparitive vs absolute assessment/evaluation

## General Disclaimer

>>>>

---
title: |
  | Infectious Disease Forecast Evaluation Based On Realized Social Utility:
  | Allocation Scores
  | ISF 43
author: Aaron Gerding, Nick Reich, Ben Rogers, Evan Ray
institute: "School of Public Health and Health Sciences, UMass, Amherst"
format:
  revealjs: 
    width: 1050
    height: 700
revealjs-plugins:
  - attribution
editor: visual
css: styles.css
---

## Introduction

::: {.column width="50%"}

About me:

- PhD in pure math
- transitioned to public health data science 
- started as RA with Nick and Evan at [reichlab.io](reichlab.io) at UMass in 2021
- maintain COVID-19 Forecast Hub ensemble
- current research goals center on ensemble methodology
  - focus on scoring rules, decision theory, forecast-policy alignment

::: {.column width="50%"}

![](Reichlab.png)

![](C19FH.png)

@@

![](hub_timeline.png){width=80%}

@@

New directions: "Hubverse" infrastructure project [https://hubdocs.readthedocs.io/](), genomic targets

@@

## Basic products of ID Forecasting Hubs

![](plots/p_VA_CU_IHME.png)

## The Role of ID forecasting Hubs

@@

- trend from point and probability (of bins) toward quantile reporting

@@

- substantial buy-in from policy makers

> It's important to note that while these efforts represent significant advances in the field, forecasting the spread and impact of infectious diseases is a very complex task with many challenges, and these models should be **interpreted with caution**.

Basic example of **interpretation** is an **allocation decision** regarding

-   medical supplies
-   facility capacity
-   personnel
-   R&D funding
-   public tolerance for NPIs

::: notes
quantiles flexible interpretable - closer to describing full distribution which DM probably always wants

multicenter effort to collect forecasts via hubs for ID outbreaks (epi/pan/endemic)

With uncertainty quantification being well-established as needed DM input,

Allocations can be wrt space, time, demographic

Next: formalize this.
:::

@@

$S$ if **proper** if $\displaystyle{E_{F}[S(F,Y)] \leq E_{F}S(G,Y)]}$ for all $F, G$

@@

Hubs have tried to use proper SRs to rank and combine forecasts in ways that offer best suppport for PH response to ID outbreaks

Primary scoring rule used currently is the **Weighted Interval Score** which is a discrete approximation to the 
**Continuous Ranked Probability Score** (a probabilistic generalization of MAE)

@@

<br>

#### A central goal in design of scoring rules for social good

::: box

@@

<br>

::: box

@@

::: box

Key point: Given $l$, we have an *automatically proper* scoring rule 
$S(F,y) := l(x^F, y)$

<br>

#### Our basic example:

@@

$$
\begin{align}
l(x,y) &= \sum_{i=1}^{52} \max(y_i - x_i, 0) = \sum (y_i - x_i)_{+}\\
x^F &\in \left\{ \text{possible } x' \,\middle| \, E_{F}[l(x',Y)] \leq E_{F}[l(x,Y)] \text{ for any possible } x \right\}
\end{align}
$$

<br>

#### Novelty of our approach for ID forecast evaluation:

::: box

Define set of **possible** $x$ as those actions which satisfy $\displaystyle{\sum x_i \leq K}$

@@

<br>

#### Note:

@@

\begin{align}
L(x, \lambda ; K, F) &= \sum_{i=1}^N E_{F_i}\left[l(x_i, Y_i)]\right)+\lambda\left(\sum_{i=1}^N x_i-K\right) \\

\nabla_{x} L(x^F, \lambda) = 0 &\implies 
E_F \left[\frac{d}{dx_i}(Y_i - x_i^F)_{+}\right] + \lambda \\
&= F_i(x_i^F)-1 + \lambda = 0 
\implies x_i^F = F_i^{-1}(1-\lambda) \\

\nabla_{\lambda} L(x^F, \lambda^{\star}) = 0 
&\implies \sum_{i=1}^{N} F_{i}^{-1}(1-\lambda^{\star}) - K  = 0
\implies 
\boxed{x_{i}^F = F_i^{-1}(1-\lambda^{\star})}
\end{align}

Our allocation scoring rule is then given by the recipe:

1. Use forecast $F$ to form Lagrangian $L$, for expected loss under a resource constraint

2. Solve for optimizer $x^F$ using $L$

3. Plug $x^F$ and outcome $y$ into loss function to get $S(F,y) = l(x^F,y)$

But getting $x^F$ not so simple.

Even for simple parametric $F_i$, there is rarely an analytic solution for $\lambda^{\star}$.

- and we need to reconstruct $F_i$'s from quantile forecasts $q_{i,j}, j = 1,\ldots,23$ to form $L(x, \lambda ; K, F)$

![](plots/dist_pqs_plot_OH.png)

## Iterative methods

So we resort to a binary search for $\lambda^{\star}$ following long history in inventory management of solving **constrained multi-product newsvendor** 
problems^[See e.g., Zhang, Bin, Xiaoyan Xu, and Zhongsheng Hua. 2009. “A Binary Solution Method for the Multi-Product
Newsboy Problem with Budget Constraint.”]

![](plots/iter_combo.png)

![](plots/nat_hosps.png)

## Results

::: {.column width="30%"}

After adjusting for an oracle the score of an oracle that allocates *fairly* based on what demands actually occur, scores resemble "value" plots from meteorology

- also the Murphy curves of Gneiting et. al.

- extreme shortages or surplusses diminish the oracle's advantage

- ranking at peak seems to be consistent with that in large $K$
region.

::: {.column width="70%"}

![](plots/peak_alloscore.png)

::: aside
Ehm, Werner, Tilmann Gneiting, Alexander Jordan, and Fabian Krüger. 2016. “Of Quantiles and Expectiles: Consistent Scoring Functions, Choquet Representations and Forecast Rankings.”
:::

::: notes
Note: The oracle is not the DGP, it is a time-traveller
:::

## Alloscore and WIS rank models differently

![](plots/p_wis_v_alloscore.png)

## Explanations?

![](plots/mvbucky_dist_alloc.png)

## Limitations

We hope/think that allocation scoring is sensitive to implicit dependence structures in forecasts, but all work so far only refers directly to marginals
- nothing yet with copulas, etc.

<br>

### Thank you!

A very rough R package I wrote to implement scoring procedures:

[https://github.com/aaronger/alloscore](https://github.com/aaronger/alloscore)

A less rough package Evan wrote to implement cdf reconstruction

[https://github.com/reichlab/distfromq](https://github.com/reichlab/distfromq)

## National picture

![](plots/state_hosps_ens_only.jpeg){width=100%}

@@

<br>

<br>

An explicit vacabulary of social loss in ID forecast evaluation has instrinsic advantage
over "off-the-shelf" metrics from statistic inference

@@

I do not have a clean definition of "social loss/good/risk/regret" but wanted to use term becasue it is in the session title

@@
